Natural variation in teosinte at the domestication locus teosinte branched1 (tb1)

### 2014-04-06 manuscript.tex
minor changes

We examined discrepancies between observed and expected genotype frequencies by calculating Hardy-Weinberg Equilibrium (HWE), and to calculate differentiation between populations (F\textsubscript{ST}) and subspecies (F\textsubscript{CT}) we used HierFstat \citep{Goudet2005}. These analyses only included populations in which 8 or more individuals were sampled. To test the hypothesis that the Hopscotch insertion may be adaptive under certain environmental conditions, we looked for significant associations between the Hopscotch frequency and environmental variables using BayEnv \citep{Coop2010}. BayEnv creates a covariance matrix of relatedness between populations, and then tests a null model that allele frequencies in populations are determined by the covariance matrix of relatedness alone against the alternative model that allele frequencies are determined by a combination of the covariance matrix and an environmental variable, producing a posterior probability (Bayes Factor)\citep{Coop2010}. We used genotyping and covariance data from Pyh\"aj\"arvi et al., (2013) \citep{Pyh\"aj\"arvi2013} for BayEnv, with the Hopscotch insertion coded as an additional SNP (\ref{TableS3BayEnv}). Environmental data were obtained from www.worldclim.org, the Harmonized World Soil Database and www.harvestchoice.org, and summarized by principle component analysis \citet{Pyh\"aj\"arvi2013} \citep{Pyh\"aj\"arvi2013}.

@@

For population genetic analyses of sequenced region 1 and sequenced region 2 we used the analysis package of Libsequence \citep{Thornton2003} to calculate pairwise F\textsubscript{ST} between populations, and to calculate standard diversity statistics (number of haplotypes; haplotype diversity; Watterson's estimator $\theta_{\subscript{W}}$; pairwise nucleotide diversity $\theta_{\subscript{\pi}}$; and Tajima's D). To produce a visual representation of differentiation between sequences and to examine patterns in sequence clustering by Hopscotch genotype we used Phylip (http://evolution.genetics.washington.edu/phylip.html) to create neighbor-joining trees with bootstrapping (100 repetitions) to examine the support of nodes in our trees. For creation of trees we also included homologous sequence data from teosinte inbred lines (TILs), some of which are known to be homozygous for the Hopscotch insertion (TIL03, TIL17, TIL09), as well as 59 lines of domesticated maize and landraces (data from Maize HapMapV2, \citep{Chia2012}).

@@

We examined evidence of introgression on chromosome 1 in these same four populations (EjuA, EjuB, MSA, SLO) using STRUCTURE \citep{Falush2003} and the same phased 55K SNP data from \citep{Pyh\"aj\"arvi2013} that we used for LD analysis, combined with the corresponding SNP data from a diverse panel of 282 maize lines \citep{Cook2012}. SNPs were anchored in a modified version of the IBM genetic map (\citep{Gerke2013}, \url{http://arxiv.org/abs/1307.7313}). We created haplotype blocks using a custom Perl script that grouped SNPs separated by less than 5kb into haplotypes. We ran STRUCTURE at K=2 under the linkage model, performing 3 replicates with an MCMC burn-in of 10,000 steps and 50,000 steps post burn-in.

@@

Using our Hopscotch genotyping data, we calculated differentiation between populations (F\textsubscript{ST}) and subspecies (F\textsubscript{CT}) for populations in which we sampled 8 or more alleles. F\textsubscript{CT} is 0 within our dataset, and we found similar levels of F\textsubscript{ST} between populations within each subspecies (0.22) and between all populations (0.23), to those reported in genome-wide estimates from previous studies \cite{Pyhajarvi2013} (\ref{Table1Fst}). 
% Table 1
\begin{table}[htbp]
  \centering
  \caption{Pairwise F\textsubscript{CT} values from sequence data}
    \begin{tabular}{rccc}
    \toprule
    \multicolumn{1}{c}{\textbf{Comparison}} & \textbf{Sequenced Region 2} & \textbf{Sequenced Region 1} & \textbf{Hopscotch} \\
    \midrule
    EjuA \& EJuB & 0     & 0     & 0 \\
    EjuA \& MSA & 0.328 & 0.326 & 0.186 \\
    EjuA \& SLO & 0.258 & 0.416 & 0.28 \\
    EjuB \& MSA & 0.365 & 0.397 & 0.188 \\
    EjuB \& SLO & 0.29  & 0.512 & 0.28 \\
    MSA \& SLO & 0     & 0.007 & 0.016 \\
    \bottomrule
    \end{tabular}
  \label{Table1Fst}
\end{table}
Although we found large variation in Hopscotch allele frequency among our populations, our BayEnv analysis did not indicate a correlation between the Hopscotch insertion and environmental variables (all Bayes Factors $<$ 1; \ref{TableS3BayEnv}).

@@

To investigate patterns of sequence diversity and linkage disequilibrium (LD) in the \emph{tb1} region, we sequenced two small ($<$1kb) regions upstream of the tb1 ORF in four populations. After alignment and singleton checking we recovered 40 and 48 segregating sites for the 50kb upstream region and the 5' UTR region, respectively. For region 1, Ejutla A has the highest values of haplotype diversity, $\theta_{\textsubscript{W}}$, and $\theta_{\textsubscript{\pi}}$, while Ejutla B and La Mesa have comparable values of these summary statistics, and San Lorenzo has much lower values. Additionally, Tajima's D is similarly negative in the two Ejutla populations and La Mesa, but is more positive in San Lorenzo (\ref{Table2Diversity}). 
% Table 2 Diversity Statistics
\begin{table}[htbp]
  \centering
  \caption{Add caption}
    \begin{tabular}{rccccc}
    \toprule
    Population & Number of Haplotypes & Haplotype Diversity & $\theta_{\textsubscript{W}}$    & $\theta_{\textsubscript{\pi}}$    & Tajima?s D \\
    \midrule
    \multicolumn{6}{c}{\textit{Sequenced region 2 (50kb upstream)}} \\
    \multicolumn{1}{c}{EJUA} & 8     & 0.89394 & 0.01548 & 0.01763 & 0.6231 \\
    \multicolumn{1}{c}{EJUB} & 8     & 0.89394 & 0.01493 & 0.01591 & 0.29504 \\
    \multicolumn{1}{c}{MSA} & 3     & 0.68182 & 0.01111 & 0.01055 & -0.22212 \\
    \multicolumn{1}{c}{SLO} & 4     & 0.74242 & 0.01167 & 0.01413 & 0.93185 \\
    \multicolumn{6}{c}{\textit{Sequenced region 1(5? UTR)}} \\
    \multicolumn{1}{c}{EJUA} & 8     & 0.85897 & 0.00874 & 0.00527 & -1.64955 \\
    \multicolumn{1}{c}{EJUB} & 5     & 0.70909 & 0.00663 & 0.00378 & -1.83123 \\
    \multicolumn{1}{c}{MSA} & 6     & 0.68182 & 0.00646 & 0.00373 & -1.75506 \\
    \multicolumn{1}{c}{SLO} & 3     & 0.31818 & 0.00176 & 0.00137 & -0.72873 \\
    \bottomrule
    \end{tabular}
  \label{Table2Diversity}
\end{table}

@@

% Table 3 $r^{2}$ values
\begin{table}[htbp]
  \centering
  \caption{$r^{2}$ values between SNPs in the \emph{tb1} region (positions 264,596,664-265,891,456) and the rest of chromosome 1, within the 5? UTR (Sequenced region 1), and within the 66,169 bp upstream region (Sequenced region 2).}
    \begin{tabular}{cccrc}
    \toprule
    \multirow{2}[2]{*}{Population} & \multirow{2}[2]{*}{Chromosome 1} & \multirow{2}[2]{*}{tb1 region} & \multicolumn{1}{c}{\multirow{2}[2]{*}{Sequenced region 1}} & \multirow{2}[2]{*}{Sequenced region 1 } \\
    \midrule
          &       &       & \multicolumn{1}{c}{} &  \\
    Ejutla A & 0.095426101 & 0.050304 & 0.747295 & 0.214933 \\
    Ejutla B & 0.068681837 & 0.051295 & 0.660354 & 0.186395 \\
    La Mesa & 0.069500533 & 0.053306 & 0.914286 & 0.766234 \\
    San Lorenzo & 0.100536784 & 0.067251 & 0.912281 & 0.636364 \\
    \bottomrule
    \end{tabular}
  \label{Table3R2}
\end{table}
 

@@

We find no evidence of recent introgression in our analyses. Clustering patterns in our NJ trees do not reflect a pattern expected if maize alleles at the \emph{tb1} locus had introgressed into populations of teosinte. We do not have sympatric maize samples from many of our populations, however, so we cannot rule out the possibility that Hopscotch alleles in these teosinte populations are segregating in sympatric maize. Moreover, analysis of linkage in the \emph{tb1} region does not reveal patterns of high LD relative to the rest of chromosome 1. The similarity in LD in the \emph{tb1} region in populations with and without the Hopscotch provides further support for an explanation other than introgression. Additionally, assignment to maize in this region in our STRUCTURE analysis is in fact lower than the average across chromosome 1 (\ref{Fig3Structure}, \ref{\Table4Q}).

 % Table 4 Assigment to maize and teosinte in tb1 region and chr 1 
\begin{table}[htbp]
  \centering
  \caption{Assignments to maize and teosinte in the \emph{tb1} and chromosome 1 regions from STRUCTURE}
    \begin{tabular}{ccccc}
    \toprule
          & \multicolumn{2}{c}{\emph{tb1} region} & \multicolumn{2}{c}{Chr 1} \\
    \midrule
    Population & Maize assignment & Teosinte assignment & Maize assignment & Teosinte assignment \\
    Ejutla A & 0.02158681 & 0.9784132 & 0.2026814 & 0.7973186 \\
    Ejutla B & 0.01888194 & 0.9811181 & 0.1872131 & 0.8127869 \\
    La Mesa & 0.9881333 & 0.0118675 & 0.8068998 & 0.1931017 \\
    San Lorenzo & 0.01551389 & 0.9844861 & 0.2048252 & 0.7951748 \\
    \bottomrule
    \end{tabular}
  \label{Table4Q}
\end{table}

@@

Variation at \emph{tb1} has also been shown to contribute to phenotypes other than tillering \citep{Clark2006}, and a recent study by \citet{StuderDoebley2012} examined the possibility of an allelic series at the \emph{tb1} locus in teosinte. \citet{StuderDoebley2012} introgressed 9 separate teosinte segments (one from \emph{Zea diploperennis}, and four from both \emph{Zea mays} ssp. \emph{mexicana} and \emph{Zea mays} ssp. {\parviglumis}) spanning the \emph{tb1} locus into an isogenic maize background and investigated their effects on previously associated phenotypes. They found that plants with teosinte chromosomal segments had greater tillering than their maize isogenic lines, and that different chromosomal segments of \emph{tb1} confer different amounts of tillering, suggesting that there are multiple genetic factors in this region that affect tillering. However, in addition to elucidating variance in tillering among \emph{tb1} teosinte segments, \citet{StuderDoebley2012} found significant variance among W22 control lines, suggesting that there are other genetic factors aside from alleles at the \emph{tb1} locus that affect tillering in maize. \citet{DoebleyStec1991, DoebleyStec1993} first attempted to map QTL controlling many of the phenotypic differences between domesticated maize and teosinte, and demonstrated the existence of numerous QTL that contribute to the differences in branching architecture between the two. Many of these loci (\emph{grassy tillers}, \emph{gt1}; \emph{tassel-replaces-upper-ears1}, \emph{tru1}; \emph{terminal ear1}, \emph{ter1}) have been shown to interact with \emph{tb1} \citep{Whipple2011, Li2012, PAPERS}, and both \emph{tru1} and \emph{ter1} have been shown to affect the same phenotypic traits as \emph{tb1} \citep{DoebleyStecGustus1995}. \emph{tassel-replaces-upper-ears1} (\emph{tru1}), for example, has been shown to act either epistatically or downstream of \emph{tb1}, affecting both branching architecture (decreased apical dominance) and tassel phenotypes (shortened tassel and shank length and reduced tassel number) \citep{Li2012}. It seems plausible that variation in some of these other loci could have affected tillering in our greenhouse population, and contributed to the lack of correlation we see between Hopscotch genotype and tillering.

@@

@@

@@

>>>>

We examined discrepancies between observed and expected genotype frequencies by calculating Hardy-Weinberg Equilibrium (HWE), and to calculate differentiation between populations (F\textsubscript{ST}) and subspecies (F\textsubscript{CT}) we used HierFstat \citep{Goudet2005}. These analyses only included populations in which 8 or more individuals were sampled. To test the hypothesis that the Hopscotch insertion may be adaptive under certain environmental conditions, we looked for significant associations between the Hopscotch frequency and environmental variables using BayEnv \citep{Coop2010}. BayEnv creates a covariance matrix of relatedness between populations, and then tests a null model that allele frequencies in populations are determined by the covariance matrix of relatedness alone against the alternative model that allele frequencies are determined by a combination of the covariance matrix and an environmental variable, producing a posterior probability (Bayes Factor)\citep{Coop2010}. We used genotyping and covariance data from Pyh\"aj\"arvi et al., (2013) \citep{Pyhajarvi2013} for BayEnv, with the Hopscotch insertion coded as an additional SNP (\ref{TableS3BayEnv}). Environmental data were obtained from www.worldclim.org, the Harmonized World Soil Database and www.harvestchoice.org, and summarized by principle component analysis \citet{Pyhajarvi2013} \citep{Pyhajarvi2013}.

@@

For population genetic analyses of sequenced region 1 and sequenced region 2 we used the analysis package of Libsequence \citep{Thornton2003} to calculate pairwise F\textsubscript{ST} between populations, and to calculate standard diversity statistics (number of haplotypes; haplotype diversity; Watterson's estimator $\theta_W$; pairwise nucleotide diversity $\theta_\pi$; and Tajima's D). To produce a visual representation of differentiation between sequences and to examine patterns in sequence clustering by Hopscotch genotype we used Phylip (http://evolution.genetics.washington.edu/phylip.html) to create neighbor-joining trees with bootstrapping (100 repetitions) to examine the support of nodes in our trees. For creation of trees we also included homologous sequence data from teosinte inbred lines (TILs), some of which are known to be homozygous for the Hopscotch insertion (TIL03, TIL17, TIL09), as well as 59 lines of domesticated maize and landraces (data from Maize HapMapV2, \citep{Chia2012}).

@@

We examined evidence of introgression on chromosome 1 in these same four populations (EjuA, EjuB, MSA, SLO) using STRUCTURE \citep{Falush2003} and the same phased 55K SNP data from \citep{Pyhajarvi2013} that we used for LD analysis, combined with the corresponding SNP data from a diverse panel of 282 maize lines \citep{Cook2012}. SNPs were anchored in a modified version of the IBM genetic map (\citep{Gerke2013}, \url{http://arxiv.org/abs/1307.7313}). We created haplotype blocks using a custom Perl script that grouped SNPs separated by less than 5kb into haplotypes. We ran STRUCTURE at K=2 under the linkage model, performing 3 replicates with an MCMC burn-in of 10,000 steps and 50,000 steps post burn-in.

@@

Using our Hopscotch genotyping data, we calculated differentiation between populations (F\textsubscript{ST}) and subspecies (F\textsubscript{CT}) for populations in which we sampled 8 or more alleles. F\textsubscript{CT} is 0 within our dataset, and we found similar levels of F\textsubscript{ST} between populations within each subspecies (0.22) and between all populations (0.23), to those reported in genome-wide estimates from previous studies \cite{Pyhajarvi2013} (\ref{Table1Fst}). 
% Table 1
\begin{table}[htbp]
  \centering
  \caption{Pairwise F\textsubscript{CT} values from sequence data}
    \begin{tabular}{rccc}
 %   \toprule
    \multicolumn{1}{c}{\textbf{Comparison}} & \textbf{Sequenced Region 2} & \textbf{Sequenced Region 1} & \textbf{Hopscotch} \\
 %   \midrule
    EjuA \& EJuB & 0     & 0     & 0 \\
    EjuA \& MSA & 0.328 & 0.326 & 0.186 \\
    EjuA \& SLO & 0.258 & 0.416 & 0.28 \\
    EjuB \& MSA & 0.365 & 0.397 & 0.188 \\
    EjuB \& SLO & 0.29  & 0.512 & 0.28 \\
    MSA \& SLO & 0     & 0.007 & 0.016 \\
%    \bottomrule
    \end{tabular}
  \label{Table1Fst}
\end{table}
Although we found large variation in Hopscotch allele frequency among our populations, our BayEnv analysis did not indicate a correlation between the Hopscotch insertion and environmental variables (all Bayes Factors $<$ 1; \ref{TableS3BayEnv}).

@@

To investigate patterns of sequence diversity and linkage disequilibrium (LD) in the \emph{tb1} region, we sequenced two small ($<$1kb) regions upstream of the tb1 ORF in four populations. After alignment and singleton checking we recovered 40 and 48 segregating sites for the 50kb upstream region and the 5' UTR region, respectively. For region 1, Ejutla A has the highest values of haplotype diversity, $\theta_W$, and $\theta_\pi$, while Ejutla B and La Mesa have comparable values of these summary statistics, and San Lorenzo has much lower values. Additionally, Tajima's D is similarly negative in the two Ejutla populations and La Mesa, but is more positive in San Lorenzo (\ref{Table2Diversity}). 
% Table 2 Diversity Statistics
\begin{table}[htbp]
  \centering
  \caption{Add caption}
    \begin{tabular}{rccccc}
%    \toprule
    Population & Number of Haplotypes & Haplotype Diversity & $\theta_W$    & $\theta_\pi$    & Tajima's D \\
%    \midrule
    \multicolumn{6}{c}{\textit{Sequenced region 2 (50kb upstream)}} \\
    \multicolumn{1}{c}{EJUA} & 8     & 0.89394 & 0.01548 & 0.01763 & 0.6231 \\
    \multicolumn{1}{c}{EJUB} & 8     & 0.89394 & 0.01493 & 0.01591 & 0.29504 \\
    \multicolumn{1}{c}{MSA} & 3     & 0.68182 & 0.01111 & 0.01055 & -0.22212 \\
    \multicolumn{1}{c}{SLO} & 4     & 0.74242 & 0.01167 & 0.01413 & 0.93185 \\
    \multicolumn{6}{c}{\textit{Sequenced region 1(5? UTR)}} \\
    \multicolumn{1}{c}{EJUA} & 8     & 0.85897 & 0.00874 & 0.00527 & -1.64955 \\
    \multicolumn{1}{c}{EJUB} & 5     & 0.70909 & 0.00663 & 0.00378 & -1.83123 \\
    \multicolumn{1}{c}{MSA} & 6     & 0.68182 & 0.00646 & 0.00373 & -1.75506 \\
    \multicolumn{1}{c}{SLO} & 3     & 0.31818 & 0.00176 & 0.00137 & -0.72873 \\
    \bottomrule
    \end{tabular}
  \label{Table2Diversity}
\end{table}

@@

% Table 3 $r^{2}$ values
\begin{table}[htbp]
  \centering
  \caption{$r^{2}$ values between SNPs in the \emph{tb1} region (positions 264,596,664-265,891,456) and the rest of chromosome 1, within the 5? UTR (Sequenced region 1), and within the 66,169 bp upstream region (Sequenced region 2).}
    \begin{tabular}{cccrc}
 %   \toprule
    \multirow{2}[2]{*}{Population} & \multirow{2}[2]{*}{Chromosome 1} & \multirow{2}[2]{*}{tb1 region} & \multicolumn{1}{c}{\multirow{2}[2]{*}{Sequenced region 1}} & \multirow{2}[2]{*}{Sequenced region 1 } \\
 %   \midrule
          &       &       & \multicolumn{1}{c}{} &  \\
    Ejutla A & 0.095426101 & 0.050304 & 0.747295 & 0.214933 \\
    Ejutla B & 0.068681837 & 0.051295 & 0.660354 & 0.186395 \\
    La Mesa & 0.069500533 & 0.053306 & 0.914286 & 0.766234 \\
    San Lorenzo & 0.100536784 & 0.067251 & 0.912281 & 0.636364 \\
    \bottomrule
    \end{tabular}
  \label{Table3R2}
\end{table}
 

@@

We find no evidence of recent introgression in our analyses. Clustering patterns in our NJ trees do not reflect a pattern expected if maize alleles at the \emph{tb1} locus had introgressed into populations of teosinte. We do not have sympatric maize samples from many of our populations, however, so we cannot rule out the possibility that Hopscotch alleles in these teosinte populations are segregating in sympatric maize. Moreover, analysis of linkage in the \emph{tb1} region does not reveal patterns of high LD relative to the rest of chromosome 1. The similarity in LD in the \emph{tb1} region in populations with and without the Hopscotch provides further support for an explanation other than introgression. Additionally, assignment to maize in this region in our STRUCTURE analysis is in fact lower than the average across chromosome 1 (\ref{Fig3Structure}, \ref{Table4Q}).

 % Table 4 Assigment to maize and teosinte in tb1 region and chr 1 
\begin{table}[htbp]
  \centering
  \caption{Assignments to maize and teosinte in the \emph{tb1} and chromosome 1 regions from STRUCTURE}
    \begin{tabular}{ccccc}
  %  \toprule
          & \multicolumn{2}{c}{\emph{tb1} region} & \multicolumn{2}{c}{Chr 1} \\
 %   \midrule
    Population & Maize assignment & Teosinte assignment & Maize assignment & Teosinte assignment \\
    Ejutla A & 0.02158681 & 0.9784132 & 0.2026814 & 0.7973186 \\
    Ejutla B & 0.01888194 & 0.9811181 & 0.1872131 & 0.8127869 \\
    La Mesa & 0.9881333 & 0.0118675 & 0.8068998 & 0.1931017 \\
    San Lorenzo & 0.01551389 & 0.9844861 & 0.2048252 & 0.7951748 \\
 %   \bottomrule
    \end{tabular}
  \label{Table4Q}
\end{table}

@@

Variation at \emph{tb1} has also been shown to contribute to phenotypes other than tillering \citep{Clark2006}, and a recent study by \citet{StuderDoebley2012} examined the possibility of an allelic series at the \emph{tb1} locus in teosinte. \citet{StuderDoebley2012} introgressed 9 separate teosinte segments (one from \emph{Zea diploperennis}, and four from both \emph{Zea mays} ssp. \emph{mexicana} and \emph{Zea mays} ssp. \emph{parviglumis}) spanning the \emph{tb1} locus into an isogenic maize background and investigated their effects on previously associated phenotypes. They found that plants with teosinte chromosomal segments had greater tillering than their maize isogenic lines, and that different chromosomal segments of \emph{tb1} confer different amounts of tillering, suggesting that there are multiple genetic factors in this region that affect tillering. However, in addition to elucidating variance in tillering among \emph{tb1} teosinte segments, \citet{StuderDoebley2012} found significant variance among W22 control lines, suggesting that there are other genetic factors aside from alleles at the \emph{tb1} locus that affect tillering in maize. \citet{DoebleyStec1991, DoebleyStec1993} first attempted to map QTL controlling many of the phenotypic differences between domesticated maize and teosinte, and demonstrated the existence of numerous QTL that contribute to the differences in branching architecture between the two. Many of these loci (\emph{grassy tillers}, \emph{gt1}; \emph{tassel-replaces-upper-ears1}, \emph{tru1}; \emph{terminal ear1}, \emph{ter1}) have been shown to interact with \emph{tb1} \citep{Whipple2011, Li2012, PAPERS}, and both \emph{tru1} and \emph{ter1} have been shown to affect the same phenotypic traits as \emph{tb1} \citep{DoebleyStecGustus1995}. \emph{tassel-replaces-upper-ears1} (\emph{tru1}), for example, has been shown to act either epistatically or downstream of \emph{tb1}, affecting both branching architecture (decreased apical dominance) and tassel phenotypes (shortened tassel and shank length and reduced tassel number) \citep{Li2012}. It seems plausible that variation in some of these other loci could have affected tillering in our greenhouse population, and contributed to the lack of correlation we see between Hopscotch genotype and tillering.

@@

%    \toprule

@@

 %   \midrule

@@

 %   \bottomrule
